---
title: "Learnr Examples"
author: "Peter Dolan"
date: "6/4/2019"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(learnr)
```

This is a chance for me to learn how to use the `learnr` package in order to make tutorials (which is what I want for DSci 3701).  Carefully **note the YAML header** (using RStudio's RMarkdown templates ensures that I don't need to memorize this):
```
title: "Learnr Examples"
author: "Peter Dolan"
date: "6/4/2019"
output: learnr::tutorial
runtime: shiny_prerendered
```

Most of this information comes from https://rstudio.github.io/learnr/index.html (some of which I typed verbatim and without local attribution).  One thing... this uses the `shiny_prerendered` engine and generates an HTML file.  If the HTML file is deleted and the document re-run the system remembers the inputs (need to figure out where that information lives).

Also, Chunks need to have unique chunk labels-- this makes it possible to store and reload student work.

**Each chunk is (without using global chunks) executed locally**.  This ensures that the exercises can be done independently.

## Running

Use 
```
learnr::run_tutorial("name",package="learnr")
```
to launch from the console.


## Devices

Types of `learnr` devices:

* narative, figures, illustrations, and equations
* code exercises (r code chunks that users can edit an execute directly) (includes options for **exercise checking**)
* Quizzes
* Videos (via youtube and vimeo)
* Interactive shiny componenets

Tutorials preserve work so users can return to the tutorial later and pick up where they left off.

## Hello Tutorial

First example of an exercise:

```{r addition, exercise=TRUE}
1+1
```

Let's see if we can use Python too:

```{python addition2, exercise=TRUE}
1+1
```
