---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Machine Learning Vocabulary and Concepts

CONFUSION MATRIX
FDR
FPR
ACCURACY
PRECISION/SENSITIVITY
RECALL
SPECIFICITY
F1
OVERFITTING
UNDERFITTING
CROSS VALIDATION
TRAINING SET
VALIDATION SET
SUPERVISED LEARNING
UNSUPERVISED LEARNING
BAGGING (BOOTSTRAP AGGREGATING)
CALSSIFICATION 
AGGREGATION LEARNING
WEAK LEARNERS
OOB error: Out of bag error (https://en.wikipedia.org/wiki/Out-of-bag_error)
BOOTSTRAPPING
BOOSTING (ESPECIALLY GRADIENT BOOSTING)
hyperparameters
GRADIENT DESCENT
REGULARIZATION(L1, L2, SPARSE REPRESENTATION)


PACKAGES:
gbm (potentially) Generalized Boosted regression models

library(rsample)      # data splitting 
library(randomForest) # basic implementation
library(ranger)       # a faster implementation of randomForest
library(caret)        # an aggregator package for performing many machine learning models
library(h2o)          # an extremely fast java-based platform

### Exercise 

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}

```

### k-fold cross validation

Before fitting a model (regression, regressoin/classification tree, neural net, whatever).  Break the data into $k$ pieces.  For each piece LEAVE IT OUT and perform the modeling.  See how good of a job the model does on the omitted data.  Do this for all $k$ pieces.  This provides a measure for how good of a job the model is doing.  The PRESS statistics and $p^2$ statistic are, essentially a form of 1-fold cross validation.


[This article](https://machinelearningmastery.com/k-fold-cross-validation/) has a good in-depth dicussion of the idea

##

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaulated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

