---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Overview {data-progressive=TRUE}

The purpose of this document is to introduce you to some basic mathematical ideas in a strongly pragmatic environment and reinforce those ideas by examples in both R and python.


The focus of this document will be on *practical applications*.  

## The math

The first mathematical idea that we are going need is **the vector**.  This is the purview of **linear algebra** and we are not going to have time to learn all of it.  The **first* key idea is as follows:

$$
\textrm{Key Idea 1: A vector is an array}
$$

This simple statement does a *terrible* disservice to the underlying mathematics, but it really captures the essence of what we, mostly, will care about.  Let's see an example:

### Vector example in R

In the language of R, what almost every other language calls an array is called a **vector**.  We can explicitly build one using the R-function `c()`, or we can use the output of a function such as `seq()` (which builds sequences of values), or we can use an expression such as `3:10`:

```{r}
vec<-c(10,20,30,40,50)
vec
vec[1] #1-based in R
vec[2]
vec[c(1,2)]
vec[c(1,1,4)]
vec[2:3]
```

### Vector example in Python

It's important to notice that in both languages we use `var[i]` to get at an individual entry, but R is 1-based and Python is 0-based.  So the **first** element in an array is `var[1]` in R and it is `var[0]` in Python^[this expression will generate an error in R].

```{python}
vec=[10,20,30,40,50]
vec
vec[0] #0-based in Python
vec[1]
indices=(1,2)
[vec[i] for i in indices]
indices=(1,1,4)
[vec[i] for i in indices]
```

### Why vectors?

If a vector is just an array... then why do we care about the mathematical idea of vectors?

This is where things get interesting.  In mathematics your first interpretation of a vector is as a list of numbers... but it comes equipped with **vector operations**.  These are things that take **two** vectors and produced a **third**.  For example:  Addition:

```{r}
vector.one<-c(10,20,30,40)
vector.two<-c(-1,2,-5,7)
vector.one+vector.two
```

```{python}
vectorOne=[10,20,30,40]
vectorTwo=[-1,2,-5,7]
vectorOne+vectorTwo
```

**DANGER**:  That did something else. That **concatenated** the two.  If we want the arrays to behave like vectors in python we need to use `numpy`:

```{python}
import numpy as np
vectorOne=np.array([10,20,30,40])
vectorTwo=np.array([-1,2,-5,7])
vectorOne+vectorTwo
```

So what makes something a vector anyway?  

### The mathematical details

stuff