---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
source("checkBasicRExercises.R") #imports ps.file
tutorial_options(exercise.checker = check_exercise)
```

## Simple Function Exercise 

Read the Week 3 material on functions.

Write R code required to take exactly 3 arguments and calculate the [geometric mean](https://en.wikipedia.org/wiki/Geometric_mean) of the three arguments.  Your function should use $x,y,$ and $z$ in the parameter list and all parameters should have a default value of 1:
```{r geometric-triplet-setup}
# The problem must be added to the problem-set in <exercise>-setup
ps$addProblem("geometric mean") #Add the exercise to the problem set 
```

```{r geometric-triplet-check}
# Checker will run original code prior to the check
ex=ps$getProblem("geometric mean")
qid="argumentValues"  #Function arguments should be x, y, and z (in that order):
ex$addQuestion(qid,correct="Correct argument names",incorrect="Incorrect argument names",pt.val=2)
if(safeTry(all(names(formals(geom.mean.3))==c("x","y","z")))){
  ex$setCorrect(qid)
}

qid="defaultValues" #Default values for the arguments should all be 1
ex$addQuestion(qid,correct="Correct default values",incorrect="Incorrect default values",pt.val=2)
if(safeTry(all(sapply(formals(geom.mean.3),function(v){v==1})))){ #This could be true even if the number of arguments is wrong, or the names are incorrect
  ex$setCorrect(qid)  
}

qid="testRun"
ex$addQuestion(qid,correct="Correct result for 2,2,-2",incorrect="Incorrect result for 2,2,-2",pt.val=2)
if(safeTry(geom.mean.3(2,2,-2)==-2)){
  ex$setCorrect(qid)  
}

qid="testRun2"
ex$addQuestion(qid,correct="Correct result for 2,3,4",incorrect="Incorrect result for 2,3,4",pt.val=2)
if(safeTry(abs(geom.mean.3(2,3,4)-2.884499)<0.001)){
  ex$setCorrect(qid)
}
exercise=ex # `exercise` holds the exercise used by the code checker
```

```{r geometric-triplet, exercise=TRUE, exercise.lines=5}
geom.mean.3<-function(...){ #replace ... with parameter list.  Your arguments should be x,y, and z
  #add body here
}
geom.mean.3(2,3,4)
```

## The if-else statement

Write a function called `parity` that uses an `if-else` statement to return the words "even" if the input is even and "odd" if the input is odd.  You may assume that the input is an integer.

```{r parity, exercise=TRUE, exercise.lines=18}

parity(0) #should return "even"
parity(3) #should return "odd"
parity(-3) #should return "odd"

```

```{r parity-setup}
# The problem must be added to the problem-set in <exercise>-setup
ps$addProblem("parity check") #Add the exercise to the problem set 
```

```{r parity-check}
# Checker will run original code prior to the check
ex=ps$getProblem("parity check")
qid="function exists"  #
ex$addQuestion(qid,correct="Function 'parity' exists",incorrect="function 'parity does not exist",pt.val=1)
if(safeTry(exists("parity"))){
  ex$setCorrect(qid)
}

qid="contains if"  #
ex$addQuestion(qid,correct="'parity' uses if()",incorrect="'parity' does not use if()",pt.val=1)
if(safeTry({grepl("if",as.character(body(parity))[2])})){
  ex$setCorrect(qid)
}

qid="contains else"  #
ex$addQuestion(qid,correct="'parity' uses else",incorrect="'parity' does not use else",pt.val=1)
if(safeTry({grepl("else",as.character(body(parity))[2])})){
  ex$setCorrect(qid)
}

qid="check 0" 
ex$addQuestion(qid,correct="correct output for 0",incorrect="incorrect output for 0",pt.val=1)
if(safeTry({parity(0)=="even"})){
  ex$setCorrect(qid)
}

qid="check -3" 
ex$addQuestion(qid,correct="correct output for -3",incorrect="incorrect output for -3",pt.val=1)
if(safeTry({parity(-3)=="odd"})){
  ex$setCorrect(qid)
}

qid="check 4" 
ex$addQuestion(qid,correct="correct output for 4",incorrect="incorrect output for 4",pt.val=1)
if(safeTry({parity(4)=="even"})){
  ex$setCorrect(qid)
}

exercise=ex
```

## A for loop

This is a two-part problem.  First you need to create a table called `zipcodeTable` and get it in the proper format
**THEN** You need to write a function .

**Table:**   

* Use `rvest` to scrape the [data on the structnet.com website](http://www.structnet.com/instructions/zip_min_max_by_state.html)  ([Here](https://www.r-bloggers.com/using-rvest-to-scrape-an-html-table/) is a nice example.)
* Call the table `zip2State`
* Clean the table:
    * Drop the ID column too.  
    * Rename the remaining columns to "State", "ST", "Zip.min", and "Zip.max" (capitalization matters)
    * Ensure Zip.min and Zip.max are numeric (or integers)

**Function:**

* Create a function called `convert` that takes two arguments:  
   * One is a **zip-code** and the other 
   * is a **table** with columns State, Zip.min, and Zip.max (it *could* have other columns too... but it needs those).  
* The output should be the state/territory(if there are no matches, then return `"**NOT FOUND**"`).
* It must use a `for` loop to test the potential zip-code range for each row in the table. (hint:  the expression `row in 1:nrow(zip2State)` might be useful to you.)
* The `zip` argument should be able to be numeric or character.

```{r state-for, exercise=TRUE, exercise.lines=23}
#Add your code here
convert(99635,zip2State) # Should return "Alaska"
convert("99635",zip2State) # So should this
```

```{r state-for-setup}
# The problem must be added to the problem-set in <exercise>-setup
ps$addProblem("for loop lookup") #Add the exercise to the problem set 
```

```{r state-for-check}
# Checker will run original code prior to the check
ex=ps$getProblem("for loop lookup")
qid="rvest"
ex$addQuestion(qid,correct="rvest is loaded",incorrect="rvest is not ",pt.val=1)
if(safeTry(any(grepl("rvest",search())))){
  ex$setCorrect(qid)
}

qid="zip2State"
ex$addQuestion(qid,correct="data.frame zip2State exists",incorrect="data.frame zip2State does not exist",pt.val=1)
if(safeTry(class(zip2State)=="data.frame")){
  ex$setCorrect(qid)
}

qid="proper columns"
ex$addQuestion(qid,correct="Proper column names in zip2State",incorrect="Improper column names (or columns) in zip2State",pt.val=2)
if(safeTry(length(colnames(zip2State)>1) & all(colnames(zip2State)==c("State","ST","Zip.min","Zip.max")))){
  ex$setCorrect(qid)
}


qid="numeric columns"  #
ex$addQuestion(qid,correct="Zip.min and Zip.max of mode numeric",incorrect="Zip.min and Zip.max not of mode numeric",pt.val=2)
if(safeTry(all(sapply(zip2State[,c("Zip.min","Zip.max")],mode)=="numeric"))){
  ex$setCorrect(qid)
}

qid="function exists"  #
ex$addQuestion(qid,correct="Function 'convert' exists",incorrect="function 'convert' does not exist",pt.val=1)
if(safeTry(exists("convert"))){
  ex$setCorrect(qid)
}


qid="numberArguments"  
ex$addQuestion(qid,correct="Correct number of arguments",incorrect="Incorrect number of arguments",pt.val=2)
if(safeTry({length(names(formals(convert))==2)})){
  ex$setCorrect(qid)
}

qid="contains for"  #
ex$addQuestion(qid,correct="'convert' uses for()",incorrect="'convert' does not use for()",pt.val=1)
if(safeTry({any(grepl("for",as.character(body(convert))))})){
  ex$setCorrect(qid)
}

qid="Check 99635"  #
ex$addQuestion(qid,correct="numeric 99635 is Alaska",incorrect="numeric 99635 is not Alaska",pt.val=1)
if(safeTry({convert(99635,zip2State)=="Alaska"})){
  ex$setCorrect(qid)
}

qid="Check '99635'"  #
ex$addQuestion(qid,correct="character 99635 is Alaska",incorrect="character 99635 is not Alaska",pt.val=1)
if(safeTry({convert('99635',zip2State)=="Alaska"})){
  ex$setCorrect(qid)
}

qid="Check 1"  #
ex$addQuestion(qid,correct="01 produces **NOT FOUND**",incorrect="01 does not produce **NOT FOUND**",pt.val=1)
if(safeTry({convert(01,zip2State)=="**NOT FOUND**"})){
  ex$setCorrect(qid)
}

exercise=ex
```


## The `apply` function

Rewrite the above function to use `apply` and NOT a for loop.  Unfortunately  you *will* need to copy-and-paste the code that prepares the `zip2State` table because each exercise is run in an isolated sandbox.

```{r state-apply, exercise=TRUE, exercise.lines=23}
#Add your code here
convert(99635,zip2State) # Should return "Alaska"
convert("99635",zip2State) # So should this 
```
```{r state-apply-setup}
# The problem must be added to the problem-set in <exercise>-setup
ps$addProblem("state lookup with apply") #Add the exercise to the problem set 
```

```{r state-apply-check}
# Checker will run original code prior to the check
ex=ps$getProblem("state lookup with apply")

qid="function exists"  #
ex$addQuestion(qid,correct="Function 'convert' exists",incorrect="function 'convert' does not exist",pt.val=1)
if(exists("convert")){
  ex$setCorrect(qid)
}

qid="numberArguments"  #Function arguments should be x, y, and z (in that order):
ex$addQuestion(qid,correct="Correct number of arguments",incorrect="Incorrect number of arguments",pt.val=2)
if(safeTry({length(names(formals(convert))==2)})){
  ex$setCorrect(qid)
}

qid="contains apply"  #
ex$addQuestion(qid,correct="'convert' uses apply()",incorrect="'convert' does not use apply()",pt.val=1)
if(safeTry({any(grepl("apply",as.character(body(convert))))})){
  ex$setCorrect(qid)
}

qid="Check 99635"  #
ex$addQuestion(qid,correct="numeric 99635 is Alaska",incorrect="numeric 99635 is not Alaska",pt.val=1)
if(safeTry({convert(99635,zip2State)=="Alaska"})){
  ex$setCorrect(qid)
}

qid="Check '99635'"  #
ex$addQuestion(qid,correct="character 99635 is Alaska",incorrect="character 99635 is not Alaska",pt.val=1)
if(safeTry({convert('99635',zip2State)=="Alaska"})){
  ex$setCorrect(qid)
}

qid="Check 1"  #
ex$addQuestion(qid,correct="01 produces **NOT FOUND**",incorrect="01 does not produce **NOT FOUND**",pt.val=1)
if(safeTry({convert(01,zip2State)=="**NOT FOUND**"})){
  ex$setCorrect(qid)
}

exercise=ex
```


## A `while` loop

Write a function called `three.n.plus.one` that takes a single argument called `n` and uses a `while()` loop to return a numeric vector.  The function should implement the following algorithm:

```
1. If `n` is not greater than 0 return `NULL` immediately
2. Round `n` up (just in case `n` is not an integer)
4. While n is not 1:
   1. If n is even divide by 2
   2. Otherwise multiply by 3 and add 1
5. Return the vector of intermediate values.
```

**Example one:** ($n=10$)   
```{r,eval=FALSE}
three.n.plus.one(10)
```

Returns

```
[1] 10  5 16  8  4  2  1
```

**Example Two:** ($n=10.5$ [secretly this $n=11$ because of the rounding])   

```{r,eval=FALSE}
three.n.plus.one(10.5)
```

Returns

```
[1] 11 34 17 52 26 13 40 20 10  5 16  8  4  2  1
```

**Example Three:** ($n=1$)   

```{r,eval=FALSE}
three.n.plus.one(1)
```

Returns

```
[1] 1
```

**Example Four:** ($n=0$)   

```{r,eval=FALSE}
three.n.plus.one(0)
```

Returns

```
NULL
```

```{r threen, exercise=TRUE, exercise.lines=25}
#Add your code here
```

```{r threen-setup}
# The problem must be added to the problem-set in <exercise>-setup
ps$addProblem("3n+1") #Add the exercise to the problem set 
```

```{r threen-check}
# Checker will run original code prior to the check
ex=ps$getProblem("3n+1")
qid="function exists"  #
ex$addQuestion(qid,correct="Function 'three.n.plus.one' exists",incorrect="function 'three.n.plus.one' does not exist",pt.val=1)
if(exists("three.n.plus.one")){
  ex$setCorrect(qid)
}

qid="contains while"  #
ex$addQuestion(qid,correct="'three.n.plus.one' uses while()",incorrect="'three.n.plus.one' does not use while()",pt.val=1)
if(safeTry({any(grepl("while",as.character(body(three.n.plus.one))))})){
  ex$setCorrect(qid)
}

qid="argumentValues"  #Function arguments should be n:
ex$addQuestion(qid,correct="Correct argument names",incorrect="Incorrect argument names",pt.val=1)
if(safeTry({all(names(formals(three.n.plus.one))==c("n"))})){
  ex$setCorrect(qid)
}

qid="Check 10"  #
ex$addQuestion(qid,correct="Input 10 generates proper output",incorrect="Input 10 does not generate proper output",pt.val=4)
if(safeTry({all(three.n.plus.one(10)==c(10,5,16,8,4,2,1))})){
  ex$setCorrect(qid)
}

qid="Check 10.5"  #
ex$addQuestion(qid,correct="Input 10.5 generates proper output",incorrect="Input 10.5 does not generate proper output",pt.val=4)
if(safeTry({all(three.n.plus.one(10.5)==c(11,34,17,52,26,13,40,20,10,5,16,8,4,2,1))})){
  ex$setCorrect(qid)
}

qid="Check 0"  #
ex$addQuestion(qid,correct="Input 0 generates proper output",incorrect="Input 0 does not generate proper output",pt.val=1)
if(safeTry({is.null(three.n.plus.one(0))})){
  ex$setCorrect(qid)
}

qid="Check 1"  #
ex$addQuestion(qid,correct="Input 1 generates proper output",incorrect="Input 1 does not generate proper output",pt.val=1)
if(safeTry({all(three.n.plus.one(1)==c(1))})){
  ex$setCorrect(qid)
}

exercise=ex
```
