---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
source("checkBasicRExercises.R",local=TRUE)
tutorial_options(exercise.checker = check_exercise)
```


## Named arguments in R

### Exercise 

Write R code required to take exactly 3 arguments and calculate the [geometric mean](https://en.wikipedia.org/wiki/Geometric_mean) of the three arguments.  Your function should use $x,y,$ and $z$ in the parameter list and all parameters should have a default value of 1:

```{r geometric-triplet-check}
cat("in here.")
```

```{r geometric-triplet, exercise=TRUE, exercise.lines=5}
geom.mean.3<-function(...){ #replace ... with parameter list.  Your arguments should be x,y, and z
  #add body here
}
geom.mean.3(2,3,4)
```

### Exercise with Code



### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

