---
title: "Math Overview"
author: "Peter Dolan"
date: "6/3/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose

The purpose of this document is to introduce you to some basic mathematical ideas in a strongly pragmatic environment and reinforce those ideas by examples in both R and python.


The focus of this document will be on *practical applications* so we'll start with a bit of computer background/review 

## Data forms

From the first course you may recall that data when it is in a form suitable for use by a computer comes in a myriad of forms. Some examples include:

* Data as table
* Data as tree
* Data as stream
* Data as image

More generally, a computer science course will introduce the interested student to a large variety of *data structures* whose purpose is to make a number of interactions (such as calculations, lookups, modifications, insertions, deletions, to name but a few) efficient.

We are not overly concerned in this course about more advanced versions of such things.  We **will** care about *data objects* which is something both R and Python can use and **arrays** (single and multi-dimensional) as well as a few structures such as `data frames` that are specific to R.  Most of the data structures that are of interest to us in Python (such as `dict`, `set`, and, `tuple`) have relatively straightforward analogs in R.

## Data types

You should also recall that computer languages have expressions known as **variables** that can hold data.  These variables contain values but they also come with a **data type**.  Sometimes these data types are simple (such as numeric values), and sometimes they are complicated (such as objects).  But one of the keys to mastering languages such as R and python lies in always being intuitively aware of the **data type** of an expression.  The data type determines what meaningful manipulations can be done with an expression.

## Data types in R

One web page that I particularly like that teaches about programming in R^[https://swcarpentry.github.io/r-novice-inflammation/] makes the distinction between **data types** and **data structures**.  In the education I received (as an undergraduate) these would **both** be examples of a data type (one would be an example of a **primitive data type** and the other **complex (or composite) data type^[https://en.wikipedia.org/wiki/Complex_data_type])**.  To make things even a bit more confusing, the R language handbook^[https://cran.r-project.org/doc/manuals/r-release/R-lang.html] uses slightly different language as well (they use the term **mode** which can be accessed via the R-function`mode()`).

Regardless, the technical distinctions add more complications than are necessary... so I've done my duty and introduced you to the terminology you might hear being used, but I am going to use **data type** and **data structure**.  So the primitive data types and data structures in R^[that we are most interested in] are:

* NULL (rare for use to encounter)
* closure (a function)
* logical (TRUEs and FALSEs)
* integer (whole numbers-- positive, negative, as well as 0)
* double (decimals)
* character (words and strings)
* list (a data structures that contains multiple values-- usually of different data types)

These data types are all values that will be reported by the use of the R function `typeof()`

Example:

```{r}
sample=list(2,2L,TRUE,2.34,"bob",typeof,NULL)
typeof(sample)
types=sapply(sample,typeof)
```

R also has the following data structures:

* vectors
* matrices
* data frames
* factors

In other languages a **vector** would be called an **array**, a **matrix** would be a multi-dimensional array, a **data frame** doesn't have any common analogs in other languages, and a **factor** would be known as an **enumerated type**.

In R **vectors** are fundamental building blocks... In R terminology a vector has a **mode**.  Think of it as generalized primitive data type.  Example include: **numeric**, **character**, **logical**, and **function**.  Note that integers and doubles would both show up as being of mode **numeric**.  While `typeof(typeof))` and `mode(typeof)` return "closure" and "function" respectively.

<!-- consider beefing up function discussion -->

<!-- add a quiz here-->


More sophisticated data can be stored as **objects** whose **class attribute** determines how the object  is treated.  For example, R uses objects for `data frames` and `Dates`:

```{r}
some.date<-as.Date("2019-09-13") #Using default format of yyyy-mm-dd
some.date
format(some.date,"%d %b %Y")

data.frame(letters=letters[1:5],nums=1:5)
```

## Data types in Python

I'll discuss Python version 3.7.3^[https://docs.python.org/3/library/datatypes.html].  We'll keep the same data type versus data structure distinction.  Some examples of data types (sub divided)

* int
* float
* str
* bool
* code objects
* null object


Python has some nice mechanisms built-in for data structures that go well beyond what we'll cover in this course.

* arrays
* datetime
* date
* dict
* list
* set
* tuple

We can use the Python-function `type()` to get the Python data type:
```{python}
type(1337)
type(True)
type(13.37)
type("stuff")
type([1,2,3])
type({'key':'value'})
type({1,2,3})
```

# The math

The first mathematical idea that we are going need is **the vector**.  This is the purview of **linear algebra** and we are not going to have time to learn all of it.  The **first* key idea is as follows:

$$
\textrm{Key Idea 1: A vector is an array}
$$

This simple statement does a *terrible* disservice to the underlying mathematics, but it really captures the essence of what we, mostly, will care about.  Let's see an example:

## Vector example in R

In the language of R, what almost every other language calls an array is called a **vector**.  We can explicitly build one using the R-function `c()`, or we can use the output of a function such as `seq()` (which builds sequences of values), or we can use an expression such as `3:10`:

```{r}
vec<-c(10,20,30,40,50)
vec
vec[1] #1-based in R
vec[2]
vec[c(1,2)]
vec[c(1,1,4)]
vec[2:3]
```

## Vector example in Python

It's important to notice that in both languages we use `var[i]` to get at an individual entry, but R is 1-based and Python is 0-based.  So the **first** element in an array is `var[1]` in R and it is `var[0]` in Python^[this expression will generate an error in R].

```{python}
vec=[10,20,30,40,50]
vec
vec[0] #0-based in Python
vec[1]
indices=(1,2)
[vec[i] for i in indices]
indices=(1,1,4)
[vec[i] for i in indices]
```

## Why vectors?

If a vector is just an array... then why do we care about the mathematical idea of vectors?

This is where things get interesting.  In mathematics your first interpretation of a vector is as a list of numbers... but it comes equipped with **vector operations**.  These are things that take **two** vectors and produced a **third**.  For example:  Addition:

```{r}
vector.one<-c(10,20,30,40)
vector.two<-c(-1,2,-5,7)
vector.one+vector.two
```

```{python}
vectorOne=[10,20,30,40]
vectorTwo=[-1,2,-5,7]
vectorOne+vectorTwo
```

**DANGER**:  That did something else. That **concatenated** the two.  If we want the arrays to behave like vectors in python we need to use `numpy`:

```{python}
import numpy as np
vectorOne=np.array([10,20,30,40])
vectorTwo=np.array([-1,2,-5,7])
vectorOne+vectorTwo
```

So what makes something a vector anyway?  

## The mathematical details

$$
\begin{aligned}
\textrm{Average Rate of Change} &= \frac{f(x+h) - f(x)}{(x+h)-x}\\
&=\frac{f(x+h) - f(x)}{h}\\
\end{aligned}
$$
IN other words it is the **change in y** over the **change in x**.  This is also known as the **slope** of a line.  You can see 