---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

```{r datatypesetup,echo=FALSE}
data.type=list(m1=data.frame(),m2=numeric(),m3=character(),m4=list(),m5=logical())
```


## Overview {data-progressive=TRUE}

The purpose of this document is to review some basic ideas about data and programming while also  introducing some new examples in both R and python.

### Data forms

From the first course you may recall that data when it is in a form suitable for use by a computer comes in a myriad of forms. Some examples include:

* Data as table
* Data as tree
* Data as graph
* Data as stream

More generally, a computer science course will introduce the interested student to a large variety of *data structures* whose purpose is to make a number of interactions (such as calculations, lookups, modifications, insertions, deletions, to name but a few) efficient.

We are not overly concerned in this course about such things.  We **will** care about *data objects* which is something both R and Python can use.

```{r data-form, echo=FALSE}
question("Which of the following best represents the type of data encoded in a JSON object?",
         answer("table"),
         answer("tree",correct=TRUE),
         answer("graph"),
         answer("stream")
)

```

### Data types

You should also recall that computer languages have expressions known as **variables** that can hold data.  These variables contain values but they also come with a **data type**.  Sometimes these data types are simple (such as numeric values), and sometimes they are complicated (such as objects).  But one of the keys to mastering languages such as R and python lies in always being intuitively aware of the **data type** of an expression.  The data type determines what meaningful manipulations can be done with an expression.

Modify the following code so that the following conditions are satisfied (you might want to review basic R syntax before proceeding)

Requirement| Condition
-----------|-----------
1          | The sum of the two numeric variables is stored in a variable called `sum`
2          | The two character variables are concatenated together into a variable called `concat`.  Let them be separated by a space when you do this.
3          | The two logical variables are `anded` together into a variable called `logical.and`
4          | The variable `df` is a data.frame with 10 rows and two columns.  The first column is named `letters` and stores the first 10 letters, while the second column is named `numbers` and contains the numbers 1 through 10
5          | Create a simple function called `myfunc` that takes the difference of two variables
```{r datatypes-check}

```

```{r datatypes, exercise=TRUE, exercise.setup="datatypesetup"}
num.1 <- 2
num.2 <- 3
sum <-  # Add code to the left of this comment
char.1 = "Data"
char.2 = "Science"
concat<- # Add code to the left of this comment
logical.1<-TRUE
logical.2<-FALSE
logical.and<- # Add code to the left of this comment
df<-data.frame(letter=,1:10)
myfunc<-function(a,b){} # Add code to the left of this comment
```

### Data types in R

One web page that I particularly like that teaches about programming in R^[https://swcarpentry.github.io/r-novice-inflammation/] makes the distinction between **data types** and **data structures**.  In the education I received (as an undergraduate) these would **both** be examples of a data type (one would be an example of a **primitive data type** and the other **complex (or composite) data type^[https://en.wikipedia.org/wiki/Complex_data_type])**.  To make things even a bit more confusing, the R language handbook^[https://cran.r-project.org/doc/manuals/r-release/R-lang.html] uses slightly different language as well (they use the term **mode** which can be accessed via the R-function`mode()`).

Regardless, the technical distinctions add more complications than are necessary... so I've done my duty and introduced you to the terminology you might hear being used, but I am going to use **data type** and **data structure**.  So the primitive data types and data structures in R^[that we are most interested in] are:

* NULL (rare for use to encounter)
* closure (a function)
* logical (TRUEs and FALSEs)
* integer (whole numbers-- positive, negative, as well as 0)
* double (decimals)
* character (words and strings)
* list (a data structures that contains multiple values-- usually of different data types)

These data types are all values that will be reported by the use of the R function `typeof()`

Example:

```{r}
sample=list(2,2L,TRUE,2.34,"bob",typeof,NULL)
typeof(sample)
types=sapply(sample,typeof)
```

R also has the following data structures:

* vectors
* matrices
* data frames
* factors

In other languages a **vector** would be called an **array**, a **matrix** would be a multi-dimensional array, a **data frame** doesn't have any common analogs in other languages, and a **factor** would be known as an **enumerated type**.

In R **vectors** are fundamental building blocks... In R terminology a vector has a **mode**.  Think of it as generalized primitive data type.  Example include: **numeric**, **character**, **logical**, and **function**.  Note that integers and doubles would both show up as being of mode **numeric**.  While `typeof(typeof))` and `mode(typeof)` return "closure" and "function" respectively.

<!-- consider beefing up function discussion -->

<!-- add a quiz here-->


More sophisticated data can be stored as **objects** whose **class attribute** determines how the object  is treated.  For example, R uses objects for `data frames` and `Dates`:

```{r}
some.date<-as.Date("2019-09-13") #Using default format of yyyy-mm-dd
some.date
format(some.date,"%d %b %Y")

data.frame(letters=letters[1:5],nums=1:5)
```

### Data types in Python

I'll discuss Python version 3.7.3^[https://docs.python.org/3/library/datatypes.html].  We'll keep the same data type versus data structure distinction.  Some examples of data types (sub divided)

* int
* float
* str
* bool
* code objects
* null object


Python has some nice mechanisms built-in for data structures that go well beyond what we'll cover in this course.

* arrays
* datetime
* date
* dict
* list
* set
* tuple

We can use the Python-function `type()` to get the Python data type:
```{python}
type(1337)
type(True)
type(13.37)
type("stuff")
type([1,2,3])
type({'key':'value'})
type({1,2,3})
```