---
title: "Multinomial Ratios"
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

The multinomial distribution provides the probability that $n$ objects can be assigned to $k$ categories.  Each category has its own associated probability so that $p_1 + p_2 + \cdots + p_k = 1$.  The binomial distribution is a special case where $k=2$.  If we let $X_i$ denote the random variable for the number of the $n$ objects found to be in the $i$th category, then in general:

$$
\textrm{P}(X_1=n_1, X_2=n_2, \ldots ,X_k=n_k) =\frac{n!}{n_1!n_2!\cdots n_k!}p_1^{n_1}p_2^{n_2}\cdots p_k^{n_k} 
$$

We are interested in the probabilities, expected values, and variances of the **ratios** of the multinomial counts such as $\frac{X_i}{X_j}$.   We shall call these values **multinomial ratios**.  Our concerns are, particularly, for when the value of $n$ is relatively small and asymptoptic considerations are not particularly accurate.

Without loss of generality we may develop our technique to apply to $X_1$ and $X_2$ (In the future we may develop the theory to include multinomial ratio probabilities conditioned upon other counts or multinomial ratios).  Call $X_1$ the numerator count and its corresponding category the numerator category.  Similarly for $X_2$ the denominator count for the denominator category.   Under these considerations, for every multinomial distribution with $k$ categories, there is a trinomial distribution corresponding to the three categories:

1. Numerator category
2. Denominator category
3. Other 

All the relevant probabilities and statistics for the multinomial ratio $\frac{X_1}{X_2}$ we be the same for this trinomial distribution.

To simplify the formulas we will use 

$$
\begin{aligned}
p&:=p_1\\
q&:=p_2\\
r&:=p_3\\
X&:=X_1\\
Y&:=X_2\\
Z&:=X_3
\end{aligned}
$$
Notice that $r=1-p-q$.

## Generating Function

The trinomial count distribution with category probabilities $p+q+r=1$ with $n$ objects has a generating function of the form

$$
f(X,Y,Z)=(pX+qY+rZ)^n
$$

Two values for $X$ and $Y$ have the same ratio $\alpha$ if 
$\frac{x}{y}=\alpha$.  In otherwords if $x=\alpha y$.  Suppose that in lowest common form $\alpha = \frac{m}{n}$.  The values of $(x,y)$ that generate the same ratio are those integers satisfying $\frac{x}{y} = \alpha$ **and** $x+y \le n$.

So... despite any other issues I can certainly write a function to calculate these values once parameters are decided upon.  For the interests of time efficiency let's make the count table and then 

```{r}
source("make.table.R")
```

This was a total accident... but it looks SOO cool I wanted to keep it:

```{r}
tmp<-make.table(100,0.4,0.4)
heatmap(tmp$count)
```

Here's the better image 

```{r}
image(tmp$count)
```

Now we are going to need to figure out which count entries are the same (and hence need to be added).  For $n=100$ the maximum value of a fraction in lowest form is 100, so the minimum distance possible between two fractions in this table are $0.01$.  So we have several interesting questions... the first is what does the density for all fractions with a max denominator of 100 and a sum of terms less than or equal to 100 look like?  Turns out that this is a VERY strange distribution because the number of count combinations that can contribute to a specific rational number will depend upon how many forms of the fraction can be built satisfying

* denominator is less than or equal to $n$
* sum of the numerator and denominator is less than or equal to $n$

```{r}
tmp<-make.table(100,0.4,0.4)
t2<-aggregate(c(tmp$count),by=list(c(tmp$index)),sum)
plot(t2$Group.1[-c(1,1421)]/256,t2$x[-c(1,1421)],xlim=c(0,2))
```

Now let's make a shiny applet to adjust the values:

```{r, echo=FALSE}
library(shiny)
fluidRow(
  column(width=3,
    sliderInput("p", "numerator probability", min = 0, max = 1, value = 0.4,step=0.05)),
column(width=3,
    sliderInput("q", "denominator probability", min = 0, max = 1, value = 0.4,step=0.05)),
column(width=3,
  sliderInput("ul", "upper limit", min = 0.5, max = 10, value = 2,step=0.5)))
sliderInput("n", "Object count", min = 10, max = 1000, value = 100,step=10)
plotOutput("density2",height="700px",width="600px")
```

```{r context="server"}
output$density2=renderPlot({
  source("make.table.R")
  n=input$n
  ul=input$ul
  tmp<-make.table(n,input$p,input$q)
  scale=attr(tmp$index,"scale")
  t2<-aggregate(c(tmp$count),by=list(c(tmp$index)),sum)
  drop=-1*c(nrow(t2)) #Drop infinity
  x=t2[drop,1]/scale
  #theta=atan2(t2[drop,1],scale)
  p=t2[drop,2]
  ev=sum(x*p)
  v=sum(p*(x-ev)^2)
  ev.txt=round(ev,3)
  sd.txt=round(sqrt(v),3)
  ratio.txt=as.character(round(input$p/input$q,3))
  msg=paste0("EV: ",ev.txt," SD:",sd.txt," ratio: ",ratio.txt)
  par(mfrow=c(2,1))
  plot(x,p,xlim=c(0,input$ul),sub=msg,main=paste0("Probability Density n=",input$n," (",input$p," , ",input$q,")"))
  abline(v=input$p/input$q)
  plot(x,cumsum(p),xlim=c(0,input$ul),main="Cunulative Distribution")
})
```

## Sanity Check

Let's calculate a few probabilities by hand and a few sanity checks to make certain that no mistakes were made.

The first one.  Since $n=100$ we can   We can, for a few fractions in lowest form determine the number of equivalent fractional forms.  For example $\frac{1}{2} = \frac{2}{4} = \frac{3}{6} = \cdots \frac{33}{66}$ so there should be $33$ indexes generated by the `make.table()` function corresponding to $\frac{1}{2}$.  Given our conventions, The index should be 1*256/2=128:

```{r}
tmp<-make.table(100,0.4,0.4)
sum(tmp$index==128)
```

The probability for this should be 

$$
\begin{aligned}
\textrm{P}\left(\frac{X_1}{X_2}=\frac{1}{2}\right)&=\textrm{P}(X_1=1,X_2=2) + \cdots + \textrm{P}(X_1=33,X_2=66)\\

&=\frac{100!}{1!2!97!}(0.4)^1(0.4)^2(0.2)^{97}+\cdots+\frac{100!}{33!66!1!}(0.4)^{33}(0.4)^2{66}0.2)^{1}\\
&=\frac{100!}{1!2!97!}(0.4)^3(0.2)^{97}+\cdots+\frac{100!}{33!66!1!}(0.4)^{99}(0.2)^{1}\\
\end{aligned}
$$

I hand-tested dmultinom to ensure that it behaved the way I think it did.  Let's make a barplot of the probabilities that contribute to 1/2:

```{r}
fracs=which(tmp$index==128,arr.ind = TRUE)
fracs=fracs-1 #because indices are 1-based but counts are 0-based
probs= apply(fracs,1,function(r){dmultinom(c(r[1],r[2],100-r[1]-r[2]),100,c(0.4,0.4,0.2))})
barplot(probs,width=1,space=0,names.arg=paste0(fracs[,1],"/",fracs[,2]))
sum(probs)
t2[t2$Group.1==128,]

```

Since infinity is a possible answer for the ratio (think 2/0) there is no mean or sd in the usual sense.

Looks like the mode is the ratio of $\frac{p}{q}$ (as it should be).  It's less clear to me what the mode is when the $\frac{p}{q}$ is not representable by an appropriate fraction.... probably the closest such fraction.

A few more notes.  Let $\mathbb{Fr}_n$ denote the space of all pairs $(p,q)$ such that $p+q \le n$.  Clearly, the relation is symmetric:  If $(p,q) \in \mathbb{Fr}_n$ then $(q,p) \in \mathbb{Fr}_n$.  

When we say $\frac{p}{q} \in mathbb{Fr}_n$ that is a shorthand for $(p,q)$.  However we will also abuse notations and let $\frac{p}{q}$ represent the set of all pairs $(p_i,q_i)$ such that $(p_i,q_i) \in \mathbb{Fr}_n$ and $p_iq=pq_i$.

Let $|\frac{p}{q}|$ denote the number of pairs $(p_i,q_i)$ such that $\frac{p_i}{q_i} = \frac{p}{q}$.  The highest such count is for the pair $(1,1)$ where $|\left\{(1,1)\right\}|=\left\lfloor\frac{n}{2}\right\rfloor$

It is also clear that every $\frac{i}{1} \in \mathbb{Fr}_n $ for $i\le n-1$ and has a unique representation for $i>\frac{n}{2}$.

The set is closed under recipricols.  We could also (as a side-path) define an operation that takes the average of the numerator and the average of the denominator  This operations does NOT descend to $\frac{p}{q}$ except when $|\frac{p}{q}|=|\frac{r}{s}|$.  Under this operation every element is idempotent.



###FAiled attempt
Further notice that the following expression will generate a polynomial in $R$ that contains the probabilities for the various ratios:

$$
\begin{aligned}
q(R) &= R^nf(R,\frac{1}{R},1)\\
&=R^n(pR+\frac{q}{R}+r)^n\\
&=R^n\left(\frac{pR^2 + rR + q}{R}\right)^n\\
&=\left(pR^2 + rR + q\right)^n
\end{aligned}
$$

## Moving to theta

More useful than the ratio, perhaps, would be the value of THETA (I'll use `atan2()`)
```{r, echo=FALSE}
library(shiny)
fluidRow(
  column(width=3,
    sliderInput("p2", "numerator probability", min = 0, max = 1, value = 0.4,step=0.05)),
column(width=3,
    sliderInput("q2", "denominator probability", min = 0, max = 1, value = 0.4,step=0.05))
)
sliderInput("n2", "Object count", min = 10, max = 1000, value = 100,step=10)
plotOutput("density",height="700px",width="600px")
```

```{r context="server"}
output$density=renderPlot({
  source("make.table.R")
  n=input$n2
  tmp<-make.table(n,input$p2,input$q2)
  scale=attr(tmp$index,"scale")
  t2<-aggregate(c(tmp$count),by=list(c(tmp$index)),sum)
  drop=-1*c(1) #We don't drop INF anymore since it converts to pi/2
  #x=t2[drop,1]/scale
  theta=atan2(t2[drop,1],scale)
  p=t2[drop,2]
  ev=sum(theta*p) #
  v=sum(p*(theta-ev)^2)
  ev.txt=round(ev,3)
  sd.txt=round(sqrt(v),3)
  ratio.txt=as.character(round(input$p2/input$q2,3))
  msg=paste0("EV: ",ev.txt," SD:",sd.txt," ratio: ",ratio.txt)
  par(mfrow=c(2,1))
  plot(theta,p,sub=msg,main=paste0("Probability Density n=",input$n2," (",input$p2," , ",input$q2,")"),col=rgb(0,0,0,1/4),pch=".",cex=2)
  abline(v=input$p2/input$q2)
  plot(theta,cumsum(p),main="Cumulative Distribution")
})
```

Odd note:  There is a "hole" in the cumulative distribution-- this is where the Expected value (after dropping infinity) of the original ratios occurs.  Worth looking into that.  Might be something profound