---
title: "Introduction to Python"
author: "Peter Dolan"
date: "9/15/2019"
output: html_document
---

```{r setup, include=FALSE}
library(reticulate)
use_python("/home/dolanp/.virtualenvs/r-reticulate/bin/python",required=TRUE)
matplotlib<-import("matplotlib")
matplotlib$use("Agg",force=TRUE) #Enable the matplotlib rendering engine.
knitr::opts_chunk$set(echo = TRUE)
```

<style>
.column-left{
  float: left;
  width: 49%;
  text-align: left;
}
.column-center{
  display: inline-block;
  width: 0%;
  text-align: center;
}
.column-right{
  float: right;
  width: 49%;
  text-align: left;
}
</style>

## Python and R

One of the *other* most popular languages in data science is Python.  We are primarily going to focus on R in this class, but you should have a basic familiarity with Python in case you end up needing to transition to using it.

The first version of Python was written by Guido van Rossum.  See [the Wikipedia article](https://en.wikipedia.org/wiki/Python_(programming_language)) for details.  Important for you is the distinction between Python 2.x and Python 3.x.  Python 2.7 was probably the most popular of the revisions in Python 2.x and has a huge code-base.  Python 3 is NOT completely back compatible with Python 2.x so there is a good chance that if you're looking up code examples online for Python you'll find something that doesn't quite work the way that you expect.  A word of warning:  `umm-rstudio2.morris.umn.edu` uses `Python 2.7.15` by default, but the JupyterHub server I installed uses one of the Python 3.x interpreters so there will definitely be differences in how things run.

I think the language does an excellent job being readable-- it looks very much like pseudo-code and uses indention rather than curly braces to indicate code blocks.  The language has a motto "there should be one--and preverably onle one-- obvious way to do it" (this is very different from R).

Python can be run via RMarkdown if the following conditions are satisfied:

* Python is installed (and correctly configured) on the RStudio server
* The library `reticulate` is installed on RStudio server

Since you are, primarily, running RStudio on the university VM, you are good to go.  And all that you need to do is make a code block with `{python}` instead of `{r}`.  

Here's an example:

```{python}
2+2
```

You can tell that the output is using a different code-engine because R would report this as 

```{r}
2+2
```

The down-side to using `{python}` code-blocks is the lack of interactivity.  For interactivity:

* You can go to the terminal and type python (use `quit()` to exit)
* Go to the Console, ensure that the `reticulate` library is loaded and use  the function `repl_python()` to start a python session in the console
* Go to the JupyterHub Server I set up and work there.

One of the cooler things about `repl_python()` is that `R` objects can be accessed using the `r` object.  You should try this now:

1. In the console assign 1337 to the variable a: `a<-1337`
2. Start the Python loop: `repl_python()`
3. Access the r value and add 3 to it: `r.a + 3`
4. Assign that to the variable b: `b=r.a+3`
5. Look at the value of b: `b`
6. Quit the REPL: `quit`

Any objects that you produce in python are then available back in R via `py$<variable>`.  

1. Try typing `py$b`

These sessions are persistant-- in other words you can go back and forth between them and they remember the changes you made.  For example, change the python variable b to 100:

1. `py$b=100`
2.  Restart the python REPL: `python_repl()`
3. Check the value: `b`
4. Quit again `quit`

As long as you have `library(reticulate)` in the **setup chunk** of your RMarkdown file you can do the same thing in your RMarkdown file:

Here's an R chunk where assign the value of 100 to `a`:

```{r}
#R code chunk
a<-100
a
```

---

Now in Python we'll add 1000 to that value. Assign the results to the variable `b` and change the R variable `a` to 10:

```{python}
#Python code chunk
b=r.a+1000
b
r.a=-10
```

---

We will verify that R can see Python's variables using `py`, that R's variable was changed, **and** we'll change the python variable `b` to -20:
```{r}
#R code chunk
py$b
a
py$b=-20
```

---
Finally we'll verify that python can see the changes:

```{python}
#Python code chunk
b
```

---

So now you can use BOTH R and Python and pass information back and forth between the two languages in a very straightforward fashion.

There is also a function called `source_python()` that will let you run Python code (in a file) from R.  It works the same way as `source()` in R, but runs Python code.  It ensures that the objects in the run-time environment are visible through `py`.  (Similar to what we saw above)

[The reticulate web page](https://rstudio.github.io/reticulate) has many, many good vignettes explaining what can be done.

## Timing Concerns

We use a fairly straightforward function in R that is very inefficient:

```{r}
Number.of.divisors=function(n){
  n=abs(floor(as.integer(n)))
  if(n==0){return(Inf)}
  count=0
  for(d in 1:n){ #This is terribly inefficient
    if(n%%d==0){count=count+1}
  }
  return(count)
}
```

This will count the number of number of divisors for a given integer $n$.

We will call it multiple times using `for`, `sapply` and `while` to get a sense of the timing:

```{r}
n=1000
n=1000 #n=10000
system.time({
  results=vector("integer",n)
  for(i in 1:n){
     results[i]=Number.of.divisors(i)
  }
})

system.time({
  results=sapply(1:n,Number.of.divisors)
})

system.time({
  results=vector("integer",n)
  i=1
  while(i<=n){
    results[i]=Number.of.divisors(i)
    i=i+1
  }
})
plot(results)
plot(table(results))
```

On `umm-rstudio2.morris.umn.edu` I ran this for three values of $n$ and measured the time in seconds:

|            | n=1,000   | n=10,000  | n=100,000
|------------|-----------|-----------|--------
|for loop    | 0.118     | 11.404    | 1247.369
|sapply      | 0.127     | 11.164    | 1463.231
|while loop  | 0.137     | 11.553    | 1109.421

We wee that the run-time's are fairly similar-- although the `while` loop seems to show more improvement at large values of $n$.

### Modeling run-time

In class I broke you up into small and groups and asked you to model the run-time-- ignoring the method. Below are some attempts to see what tranformations on $x=n$ and $y=\textrm{run time}$ made analysis simple:

```{r}
x=c(rep(1000,3),rep(10000,3),rep(100000,3))
y=c(0.118,0.127,0.137,11.404,11.164,11.553,1247.369,1463.231,1109.421)
plot(x,y)
x.log=log10(x)
y.log=log10(y)
plot(x,y.log)
plot(x.log,y.log)
plot(x.log,y.log) #the Best

model<-lm(y.log~x.log)
abline(model)
summary(model)
```

From the summary we see that 

$$
\log_{10}(y) = -6.90862 + 1.99900\log_{10}(x)
$$
Applying $\lambda \mapsto 10^\lambda$ to both sides of the equation, rounding and simplifying:

$$
y=\frac{1}{10,000,000}x^2
$$

From this we see that the run-time of the equation scales quadratically-- this is in keeping with a more thorough analysis.

### Bringing in Python

Now we bring in Python and do some timing:

```{python}
import math
from timeit import default_timer as timer
def numberOfDivisors(n):
    n=abs(int(n))
    if n==0:
       return math.inf
    count=0
    for d in range(1,n):
      if n%d == 0:
        count=count+1
    return count
    
numberOfDivisors(100)
n=1000
start=timer()
results=[0 for x in range(n)]
for i in range(1,n):
  results[i-1]=numberOfDivisors(i)
end=timer()
print(end-start)
```

Running the same function in Python for the same values of $n$ **on the same virtual machine** (the value produced above will be different because this is not being knitted on umm-rstudio2):

|          | 1,000   | 10,000  | 100,000   |
|----------|---------|---------|-----------|
|Python    | 0.06101 | 3.38664 | 331.35065 |

Let's compare.

```{r}
plot(x.log,y.log,ylim=c(-2,3)) 
abline(model)
points(3:5,log10(c(0.06101,3.38664,331.35064)),col="blue",cex=2,pch=20)
```

We actually see similar scaling, but the slope of the line is a bit different.

## Python Configuration

A computer can have multiple versions of Python installed.  There are also a myriad of ways in which packages can be updated **and** different configurations can be made to coexist on the same computer (things like **virtual environments**).  Read [this excellent blog post](https://opensource.com/article/19/4/managing-python-packages) on managing python packages.  Be prepared to ask questions about it in class on Wednesday.  I am not an expert in configuring Python either so my goal in that conversation is just to ensure that we all have a rough idea of what options are out there.  Key package managers are:

1. The one that comes with your OS and only tangentially are related to Python (Windows .msi files, the `dnf` utility on Fedora linux (or `yum` on older Fedora variants), the `apt` utility on Debian variants of linux, `pacman` on archlinux variants and `brew` (which is kinda-sorta official on Mac)
1. The `pip` utility
1. The `conda` utility

Each version of python can also have multiple `conda` environments and multple virtual environments.  For your own personal use you might be using a number of these, but in many real-world situations somebody else will be in charge of maintaining yoru environment.

For us, we will use the `py_discover_config()` utility in `reticulate` to discover what RStudio can figure out about installed python versions:

```{r,eval=FALSE}
py_discover_config()
```

This is what I observed:

```
python:         /usr/bin/python
libpython:      /usr/lib64/python2.7/config/libpython2.7.so
pythonhome:     /usr:/usr
version:        2.7.15 (default, Oct 15 2018, 15:24:06)  [GCC 8.1.1 20180712 (Red Hat 8.1.1-5)]
numpy:          /usr/lib64/python2.7/site-packages/numpy
numpy_version:  1.14.5

python versions found: 
 /usr/bin/python
 /usr/bin/python3
```

We see that the current default configuration is 2.7.15, but there **is** a python3 version installed (see above).

To switch to that version of Python we'll need to use the following command:  `use_python("/usr/bin/python3",required=TRUE)`.  If we want to use that version of Python in the console we'll need to restart R, but it's not a big deal if we do this in our setup chunk at the beginning of an RMarkdown file.

When my startup block looks like this:

```{r eval=FALSE}
library(reticulate)
use_python("/usr/bin/python3",required=TRUE)
knitr::opts_chunk$set(echo = TRUE)
```

Then I see something different when I run `py_discover_config`:

```{r,eval=FALSE}
py_discover_config()
```

```
## python:         /usr/bin/python3
## libpython:      /usr/lib64/libpython3.6m.so.1.0
## pythonhome:     /usr:/usr
## version:        3.6.8 (default, Mar 21 2019, 10:08:12)  [GCC 8.3.1 20190223 (Red Hat 8.3.1-2)]
## numpy:           [NOT FOUND]
## 
## NOTE: Python version was forced by use_python function
```

So you can force RMarkdown to use either version of Python... **however** we do NOT have numpy installed on umm-rstudio2.morris.umn.edu.  And we really do want it... so here's how I would install it:

```{r,eval=FALSE}
py_install("numpy")
```

Here is what I observed:

```
> py_install("numpy")
Creating virtual environment '~/.virtualenvs/r-reticulate' ...
Using python: /usr/bin/python3.6
Collecting pip
  Cache entry deserialization failed, entry ignored
  Downloading https://files.pythonhosted.org/packages/30/db/9e38760b32e3e7f40cce46dd5fb107b8c73840df38f0046d8e6514e675a1/pip-19.2.3-py2.py3-none-any.whl (1.4MB)
Collecting wheel
  Downloading https://files.pythonhosted.org/packages/00/83/b4a77d044e78ad1a45610eb88f745be2fd2c6d658f9798a15e384b7d57c9/wheel-0.33.6-py2.py3-none-any.whl
Collecting setuptools
  Downloading https://files.pythonhosted.org/packages/b2/86/095d2f7829badc207c893dd4ac767e871f6cd547145df797ea26baea4e2e/setuptools-41.2.0-py2.py3-none-any.whl (576kB)
Installing collected packages: pip, wheel, setuptools
Successfully installed pip-19.2.3 setuptools-41.2.0 wheel-0.33.6
Using virtual environment '~/.virtualenvs/r-reticulate' ...
Collecting numpy
  Downloading https://files.pythonhosted.org/packages/e5/e6/c3fdc53aed9fa19d6ff3abf97dfad768ae3afce1b7431f7500000816bda5/numpy-1.17.2-cp36-cp36m-manylinux1_x86_64.whl (20.4MB)
Installing collected packages: numpy
Successfully installed numpy-1.17.2
```

This produces a new virtual environment:

```{r,eval=FALSE}
py_discover_config()
```

```
> py_discover_config()
python:         /home/dolanp/.virtualenvs/r-reticulate/bin/python
libpython:      /usr/lib64/libpython3.6m.so.1.0
pythonhome:     /usr:/usr
version:        3.6.8 (default, Mar 21 2019, 10:08:12)  [GCC 8.3.1 20190223 (Red Hat 8.3.1-2)]
numpy:          /home/dolanp/.virtualenvs/r-reticulate/lib64/python3.6/site-packages/numpy
numpy_version:  1.17.2

python versions found: 
 /home/dolanp/.virtualenvs/r-reticulate/bin/python
 /usr/bin/python
 /usr/bin/python3
```

HOWEVER!  `numpy` will not be found if I try to use it in RMarkdown with the default header.  I need to change my `setup` block to look like this:

```{r eval=FALSE}
library(reticulate)
use_python("/home/dolanp/.virtualenvs/r-reticulate/bin/python",required=TRUE)
knitr::opts_chunk$set(echo = TRUE)
```

**NOW** I get the following configuration:

```{r}
py_discover_config()
```

All is well.  

## Recap

* The R library `reticulate` allows R sessions and python sessions to share their objects pretty seamlessly
* The `repl_python()` function switches to python in the RStudio console
* `source_python()` allows python code (stored in files) to be run from R
* `py_discover_config()` and `use_python("<dir>",required=TRUE)` allow different versions of Python and different environments to be used both in the console and in RMarkdown.

